\section{Compound Interest} \index{interest!compound} \label{sec:CompoundInterest}
\ifsolns Student page {105} \fi


\begin{enumerate}

  \item \boxedblank[2in]{\textbf{Compound Interest Formula:} 
	\ifsolns $F=P(1+\frac{r}{n})^{nt}$ \par
	If desired, let $i=\frac{r}{n}, m=nt$, then $F=P(1+i)^{m}$\fi
	} \index{interest!compound}

  \item Notes on \defnstyle{nominal interest rate} and \defnstyle{interest per period}:\fillwithlines{\stretch{1}}\index{interest!nominal} \index{interest!per period}

  \item Suppose \$1000 is deposited into an account that yields 9\% annually.
        Find the amount in the account at the end of the fifth year if the compounding is done
        \begin{enumerate}
          \item annually \index{annually}\solution{\$1538.62}
                \medskip
          \item semiannually \index{semiannually} \solution{\$1552.97}
                \medskip
          \item quarterly\index{quarterly} \solution{\$1560.51}
                \medskip
          \item monthly\index{monthly} \solution{\$1565.68}
                \medskip
          \item weekly\index{weekly} \solution{\$1567.70}
                \medskip
          \item daily\index{daily} \solution{\$1568.23}
                \medskip
        \end{enumerate}
        %\vspace{-1in}

\clearpage

  \item You take out a short-term business loan of \$5000 for a period of 5 months.
        The interest rate is 4\%, compounded monthly.
        How much will you have to repay at the end of the loan? \solution{\$5083.89}
        \vfill
  
  \item You deposit \$538 in a savings account that earns 3\% interest, compounded monthly.
        How much interest will you earn in four years?  (Your answer should be in dollars.) \solution{\$68.50}
        \vfill

  \item \setcounter{tempyear}{\arabic{currentyear}-17}
        In January \arabic{tempyear} Zeke's uncle deposited some money into a college savings account for Zeke.
        The account paid 5\% interest, compounded quarterly.
        In January \arabic{currentyear} there was \$28,626.40 in the account.
        How much did Zeke's uncle deposit back in \arabic{tempyear}? \solution{\$12299.98}

        \vfill
  %
  \item Steve takes out a loan of \$10,000.00 for four years.
        At the end of the four years, he pays back \$11,273.30 to the bank.
        If the interest was compounded monthly,
        what was the interest rate? \solution{3.00\%}
        \vfill
        
%%%%   \clearpage
%%%%
%%%%     \item Moose Jaw State Bank offers an savings account paying 5.4\% interest, compounded daily.
%%%%           The savings account at Moose Lake Credit Union pays 5.41\% interest, compounded quarterly.
%%%%           Finally, Moose Kneecap Savings \& Loan offers a CD earning 5.5\% interest, compounded annually.
%%%%           You have \$4500 to invest.
%%%%           \begin{enumerate}
%%%%             \item If you deposit your money in Moose Jaw State Bank, how many dollars of interest will you \emph{really} earn in one year?
%%%%                   What percent is that of your deposit?
%%%%                   \vfill
%%%%             \item If you instead deposit your money at Moose Lake Credit Union, 
%%%%                   how many dollars of interest will you have earned one year later?
%%%%                   What percent is that of your deposit?
%%%%                   \vfill
%%%%             \item Finally, how many dollars of interest would you earn if you put your money into Moose Kneecap Savings \& Loan for one year?
%%%%                   What percent is that of your deposit?
%%%%                   \vfill
%%%%             \item Which account is best?
%%%%                   \vfill
%%%%           \end{enumerate}
%%%%           \vspace{-1in}
%%%%
   \clearpage

   \item \boxedblank{\textbf{Annual Effective Yield Formula\footnote{Sometimes this is referred to as Annual Percentage Rate (APR)}:}
	\ifsolns $APR= (1+\frac{r}{n})^n-1$ \par
	or $APR= (1+i)^n-1$\fi}
 
\item If you open a savings account at Muddy Lake Credit Union, 
      you will earn 3\% interest compounded daily.
      What is the annual effective yield?
      \solution{ANSWER: 3.0453\%}
      \vfill

\item If you open a CD at Muddy Ridge State Bank,
      you will earn 3.2\% interest compounded quarterly.
      What is the annual effective yield?
      \solution{ANSWER: 3.2386\%}
      \vfill

\item Your financial adviser offers you two investment funds.
      If you invest in the AmerEquiShare Fund, you can expect a return of 9\% compounded semiannually (twice a year).
      If you invest in the PreferredSource Fund, you can expect a return of 8.8\% compounded monthly.
      \begin{enumerate}
        \item What is the annual effective yield for the AmerEquiShare Fund? \solution{9.20\%}
      \vfill
        \item What is the annual effective yield for the PreferredSource Fund? \solution{9.16\%}
      \vfill
        \item Which investment offers the higher return?\vfill
      \end{enumerate}

\end{enumerate}

%</WORKSHEETS>

%<*HWHEADER>
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\HOMEWORK
%</HWHEADER>

%<*HOMEWORK>


\begin{Fenumerate}

  \item Given the principal $P$, the interest rate $r$, the time $t$, and the frequency of compounding,
        find the future amount $F$.
        \begin{enumerate}
          \def\prct#1#2#3#4{$P={}$\$#1, $r=#2\%$ compounded #3, $t={}$#4}
          \item \prct{15,000}{6.5}{quarterly}{2 years}
                \ifsolns \par\soln \fbox{\$17,064.58} \else \vfill \fi
                \studentsoln{\$17,064.58}
          \item \prct{100,000}{3}{monthly}{30 years}
                \ifsolns \par\soln \fbox{\$245,684.22} \else \vfill \fi
          \item \prct{1,500}{4}{daily}{60 days}
                \ifsolns \par\soln \fbox{\$1,509.89} \else \vfill \fi
                \studentsoln{\$1,509.89}
          \item \prct{6,000}{5}{monthly}{6 months}
                \ifsolns \par\soln \fbox{\$6,151.57} \else \vfill \fi
        \end{enumerate}
%\vfill
        %Solve for future value
  \item When Susan Schroeckenthaler was born, her parents deposited \$4,000 into 
        a bank account bearing 5\% interest, compounded monthly.
        When she reached age 18, how much was in her account?
        \ifsolns
          \soln \fbox{\$9,820.03}
        \fi
        \studentsoln{\$9,820.03}
  \vfill
	\hwnewpage
        %Solve for interest 
  \item Seventh National Bank is lending \$10,000 to Frank Miller
        for three years.
        The bank compounds interest quarterly.
        If the bank needs to receive \$2,300 in interest from Frank
        to cover its expenses,
        what interest rate should it charge?
        \ifsolns
          \par\soln \fbox{6.96\% interest.}
        \fi
        \studentsoln{6.96\%}
  \vfill
        %Solve for initial deposit
  \item Stephanie deposited some money in a bank account earning 3\% interest, compounded daily.
        Twelve years later, there was \$1,720 in the account.  
        How much money did Stephanie originally deposit?
        \ifsolns
          \soln \fbox{\$1200.02}
        \fi
        \studentsoln{\$1200.02}
\vfill
  \item How much money must Igor Savelyev deposit today
        in order to have \$50,000 in twenty years, if his account bears
        \begin{enumerate}
          \item 8\% interest, compounded annually?
                \ifsolns \soln \fbox{\$10,727.41} \else \vfill \fi
                \studentsoln{\$10,727.41}
          \item 8\% interest, compounded quarterly?
                \ifsolns \soln \fbox{\$10,255.49} \else \vfill \fi
          \item 8\% interest, compounded monthly?
                \solution*{ \fbox{\$10,148.57}} \vfill
          \item 8\% interest, compounded daily?
                \ifsolns \soln \fbox{\$10,096.60} \else \vfill \fi
        \end{enumerate}
%\vfill
\end{Fenumerate} \ENDHOMEWORK
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage