\section{Study Guide} \label{sec:DivsionStudyGuide}


To prepare for the exam over this chapter,
you should review the in-class worksheets and homework.
Be ready to do the kind of problems you faced on the homework.

As a general guide, I recommend reviewing the following topics.

\begin{enumerate}
  \item Know what it means for a division of assets to be fair.
  \item Calculate the fair division by method of
        \begin{enumerate}
          \item Divider-Chooser
          \item Lone-Divider
          \item Sealed Bids
        \end{enumerate}
  \item Know what to do when two players want one and only one piece of the division.
  \item Apportionment
        \begin{enumerate}
          \item Know how to find the Standard Divisor and the Standard Quota.
          \item Understand the quota rule.
   				\item Be able to understand when a modification of the apportionment problem violates the Population Paradox.
   				\item Be able to understand when a modification of the apportionment problem violates the New States Paradox.
   				\item Be able to understand when a modification of the apportionment problem violates the Alabama Paradox.
        \end{enumerate}
   \item Given an apportionment problem, be able to find the apportionment using Hamilton's Method.
    \item Given an apportionment problem, be able to find the apportionment using Jefferson's Method.
    \item Given an apportionment problem, be able to find the apportionment using Adam's Method.
    \item Given an apportionment problem, be able to find the apportionment using Webster's Method
\end{enumerate}
