{\large Worksheet for calculating Jefferson's Method:}

\begin{enumerate}
	\item What is the total population? \label{parta} \hrulefill
	\item How many seats are you apportioning\footnote{If you aren't given this number, it is the total of the Standard Quotas}? \label{partb} \hrulefill
	\item Calculate the Standard Divisor (divide your population by the number of seats):  \hrulefill
	\item Calculate all the Standard Quotas (Q.) (divide the population of each state by your Standard Divisor).  The total of your standard quotas should be the same as the number of seats.  If that is not the case, you have rounded off too much.:
	
	\begin{center}
			\begin{tabular}{l|c|c|c|c|c|c|c|p{36pt}} \hline
	State	&	Pop. &Q. &	LQ&  MLQ	&MLQ & MLQ 	&  MLQ &	 	Appor\-tionment \\\hline
	Divisor&--&&--&&&&&\\\hline
\raisebox{0pt}[72pt][72pt]{\makebox[36pt]{}}&\makebox[36pt]{}&\makebox[36pt]{}&\makebox[36pt]{}&\makebox[36pt]{}&\makebox[36pt]{}&\makebox[36pt]{}&\makebox[36pt]{}\\ \hline
Totals &&&&&&&\\
		\end{tabular}
	\end{center}
	\item Write down all the Lower Quotas (LQ) by cutting off the decimal places.  The total of your lower quotas should be less than the number of seats.
	\item Guess a new, Modified Divisor which is less than, but close to, your Standard Divisor.  Use that to calculate new Modified Lower Quotas.  If the total of your MLQ is the same as the number of seats, you are finished.  Otherwise, you need to repeat this process until you find a MD that gives you the correct number of seats.
\end{enumerate} 