\section{Mortgages and Equity} \label{sec:MortagesEquity}

\ifsolns Student page {125} \fi
\index{equity}\index{collateral}\index{down payment}
\begin{enumerate}
  \item \defnstyle{Collateral} for a loan is \ldots \fillwithlines{\stretch{1}}
  \item A \defnstyle{mortgage} is \ldots \fillwithlines{\stretch{1}}
  \item A \defnstyle{down payment} is \ldots \fillwithlines{\stretch{1}}
  \item You want to buy a \$160,000 house.  Your bank requires a 17\% down payment.
        What will your down payment be?\solution{\$27,200.00}
        \vfill
  \item You have \$11,000 saved up for a down payment on a house.
        What is the most expensive house you can buy if your
        credit union requires a 20\% down payment? \solution{\$55,000.00}
        \vfill
  \pagebreak
  \item Guinevere de Lubac owns a house but still owes \$103,000 on her mortgage.
        She suddenly develops a horrible disease and needs to get cash, fast, for the medical treatments.
        If she can sell her house for \$237,900, how much cash will she have available? \solution{\$134,900.00}
        \vfill
  \item In a house, \defnstyle{equity} means \ldots\fillwithlines{\stretch{1}}
  \item \defnstyle{Equity} can always be calculated by\ldots\solution{Value of House minus money left on mortgage}\fillwithlines{\stretch{1}}
  \item Steve's monthly mortgage payment is \$836.25, his interest rate is 5\% compounded monthly,
        and he has 106 monthly payments left to make.
        How much money does Steve still owe on his loan? \solution{\$71,538.68}
        \vfill
  \clearpage
  \item John Johnson bought his house for \$225,000 by getting a 30-year loan at 4.5\% interest compounded monthly. His down payment brought the  monthly mortgage payment to \$1,013.37.
        Today he only has 15 years left on his loan.
        Assuming his home has not changed in value, how much equity does he have in his house? \solution{\$92,532.17}
        \vfill
  \item Mark Steubenfeld bought his house twenty years ago for \$150,000, using a 20\% down payment
        and a 30-year mortgage at 6\% interest.  His monthly payment is \$719.46.
        \begin{enumerate}
          \item How much money does Mark still owe on his loan? \solution{\$64,804.25}
                \vfill
          \item If Mark's house has increased in value 55\% over the past twenty years,
                how much equity does he have in his house now? \solution{\$167,695.75}
                \vfill
        \end{enumerate}
  \clearpage
  \item Julie and Dan Flaten bought a house in 2006 for \$200,000 with a 5\% down payment.
        Their 25-year mortgage had a 5\% interest rate.
        By 2010, however, thanks to the 2008 housing crash, their home's value had unfortunately decreased by 18\%.
        \begin{enumerate}
					\item What is their monthly payment?\vfill
					\item How much do they still owe on the house?\vfill
          \item How much equity do Julie and Dan now have in their home? \solution{Monthly Payment is \$1,110.72.  They owe \$173,085.37 on the house.  Therefore, they owe \$9,085.37 more than the house is worth.
}
                \vfill
          \item If Dan's company sends him to another city and they have to sell their home, what will happen?
                \fillwithlines{\stretch{1}}
        \end{enumerate}
\end{enumerate}

%</WORKSHEETS>

%<*HWHEADER>
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\HOMEWORK
%</HWHEADER>

%<*HOMEWORK>


\begin{Fenumerate}
  \item Ted and Margery Kolstad have saved up \$16,000 for a down payment on a new house.
        If their bank needs them to make a 15\% down payment, what is the most expensive house they could get?
        \solution*{\$106,666.67.}
  \vfill \item Erik and Annabelle Durling have \$13,000 saved for a down payment on their first house.
        They have looked at their budget and figured they can spend a total of \$975 each month on their mortgage payment.
        They can get a 30-year loan at 5\% interest.
        What is the most expensive house they can afford according to these criteria?
        \solution{\$194,624.58.}
  \vfill \item Gerhard and Maria Klaus bought a house sixteen years ago on a 30-year mortgage at 5\% interest.
        Their monthly mortgage payment is \$987.13.
        \begin{enumerate}
           \item How much do they still owe on their loan?
                \solution{\$119,093.36}
          \vfill \item If their house is worth \$287,000 today, how much equity do they have?
                \solution*{\$167,906.64}\vfill
        \end{enumerate}
\hwnewpage
  \item \setcounter{tempyear}{\arabic{currentyear}-18}
        Clarence Phelps bought a house in \arabic{tempyear} for \$158,000.
        At that time he put 15\% down and got a 25-year mortgage at 4.75\% interest.
        Since \arabic{tempyear} his home's value has increased by 60\%.
        What is Clarence's equity in his house today (in the year \arabic{currentyear})?
        \solution*{
          We first find that his mortgage payment is \$765.67.
          Thus he still owes \$54,625.41 on the loan.
          His home's value is now \$252,800, so his equity is
          \fbox{\$198,174.59.}
        }
  \vfill \item \setcounter{tempyear}{\arabic{currentyear}-12}
        Martin and Sue Brinkley bought a house in \arabic{tempyear} for \$173,000.
        They had a 20\% down payment and financed the rest with a 30-year mortgage at 6.5\% interest.
        Since that time their home's value has decreased by 12\%.
        What is the Brinkleys' equity in their house today (in the year \arabic{currentyear})?
        \solution{\$41,023.98}
				
\hwnewpage
  \vfill \item Kasumi Yakamoto has made three year's worth of payments on her 20-year mortgage.
        She originally borrowed \$110,000 and her interest rate is 6\%.
        She now needs to sell her house and move to Cleveland to pursue her dream of working at the Rock \& Roll Hall of Fame.
        How much does she need to sell her house for in order not to be in debt?
        \solution{\$100,635.09}
  \vfill \item A \defnstyle{home equity loan}, sometimes called a \defnstyle{second mortgage},
        lets people borrow money using as collateral their equity in their houses.
        Brian Akins has a home currently worth \$119,000 and 17 years of loan payments left;
        his monthly payment is \$629.25 and his interest rate is 8\%.
        He needs to do some major repairs on his house, so he approaches Smithson Bank about a home equity loan.
        Smithson Bank is willing give him a home equity loan of up to 90\% of the equity he has in his house.
        What is the most money he could borrow for the repairs?
				\ifsolns 
				\[ 629.25\frac{1-(1+.08/12)^{-204}}{.08/12}=\$70,052.35\]
				\[\$119,000-\$70,052.35=\$48,947.65\]
				\[\$48,947.65*90\% = \$44,052.89 \text{ that could be borrowed}\]
				\fi
        \solution*{\$44,052.89} \vfill
\end{Fenumerate} \ENDHOMEWORK
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%<*WORKSHEETS>

\clearpage