\subsection{Method of Sealed Bids} \label{sec:SealedBids}
\index{sealed bids}
Sometimes you want to split up items without physically cutting them apart.  For example, suppose Sally is living with two roommates and everyone is graduating.  Over the years, they have purchased some appliances together and now they need to fairly divide the appliances.  

In the method of sealed bids, each roommate is going to write down what they believe is a fair value for each appliance and seal the ``bids'' in an envelope.  When the envelopes are opened, we get the following results:

\begin{tabular}{llll}
 &  Rachel  &  Sally  &  Theresa \\\hline
 T.V.  &  \$75  &  \$100  &  \$120 \\\hline
 Microwave  &  \$35  &  \$30  &  \$40 \\\hline
 Crock Pot  &  \$5  &  \$10  &  \$7 \\\hline
 Toaster Oven  &  \$7  &  \$10  &  \$5 \\\hline
 Blender  &  \$20  &  \$25 &  \$20 \\\hline\ifsolns
 Total &\$142	&\$175	&\$192 \\\hline
Fair Share &\$ 47.33	&\$	58.33&\$ 64.00\\\hline
 \else
 Total &\$	&\$	&\$ \\\hline
Fair Share &\$	&\$	&\$ \\\hline
 \fi
 
\end{tabular}
\begin{enumerate}
	\item How much does each person think the whole property to be split is worth?  What would be their fair share? 
	\ifsolns  \$47.33 	 \$58.33 	 \$64.00 
\else
	\fillwithlines{\stretch{1}}
\fi
	
	\item How would you propose splitting up the property?
\fillwithlines{\stretch{1}}
	%\vfill
	
	\clearpage
	
	\item Method of Sealed Bids
	\begin{enumerate}
	\item Assign each item to the highest bidder.  Who gets what?
	\begin{center}
	\large
	\begin{tabular}{lp{3in}}
	Roommate & Items  \\\hline \ifsolns
	Rachel & \\\hline
	 Sally &   Crock Pot, 
 Toaster Oven, 
 Blender 
\\\hline
	 Theresa & T.V., 
 Microwave 
\\\hline
\else
		Rachel & \\\hline
	 Sally &  \\\hline
	 Theresa &\\\hline
\fi
	\end{tabular}
\end{center}
\normalsize
	\item What is the total value of the items to the winners?  Use the prices they have assigned themselves.
	\begin{center}
	\large
		\begin{tabular}{lcccc}
	Roommate & Value  & Fair Share & Cash Paid & Cash Received\\\hline
\ifsolns
Rachel  &  &  \$47.33  &    &  \$47.33 \\\hline
 Sally  &  \$45.00  &  \$58.33  &    &  \$13.33 \\\hline
 Theresa  &  \$160.00  &  \$64.00  &  \$96.00  &    \\\hline
\else
	Rachel & \$ \\\hline
	 Sally & \$ \\\hline
	 Theresa & \$ \\\hline
	\fi
	\end{tabular}
\normalsize
\end{center}	 
	 \item Which of the roommates have received more than their fair share of the property?  How much more did they receive? \label{sb1}

	 \item Which of the roommates have received less than their fair share of the property?  How much less did they receive? \label{sb2}
	\item Have the roommates who received too much pay cash for the extra value of the property over and above their fair share.  Pay the roommates who did not receive enough the shortfall between what they see as a fair share and the value of the property they have received.
	\item How much cash do you have left? \solution{ \$35.33} 
	\item Divide the extra cash by 3 and give each roommate one third of the extra cash.
	\ifsolns
	\begin{tabular}{lcr}
Roommate  & \\
Rachel  &  \$59.11 \\
 Sally  &  \$25.11  & Crock Pot,  Toaster Oven,  Blender\\
 Theresa  &  \$(84.22) & T.V.,  Microwave\\
\end{tabular} \else
	\fillwithlines{\stretch{1}}
\fi
	

\item Explain how each roommate has received a fair share.
	\fillwithlines{\stretch{1}}
	

\item What are some of the drawbacks of the  method of sealed bids?

\fillwithlines{\stretch{1}}
\end{enumerate}


\clearpage

\item Five heirs ($A,B,C, D,$ and $E$) are dividing an estate consisting of six items using the method of sealed bids.  The heirs bids on each of the times are given in the following table.

\begin{center}
	\begin{tabular}{c c c c c c}
	& $A$ & $B$ & $C$ & $D$ & $E$ \\\hline
	\ifsolns
	Item 1  &  \$352.00  &  \$295.00  &  \$395.00  &  \$368.00  &  \$324.00 \\\hline
Item 2  &  \$98.00  &  \$102.00  &  \$98.00  &  \$95.00  &  \$105.00 \\\hline
Item 3  &  \$460.00  &  \$449.00  &  \$510.00  &  \$501.00  &  \$476.00 \\\hline
Item 4  &  \$852.00  &  \$825.00  &  \$832.00  &  \$817.00  &  \$843.00 \\\hline
Item 5  &  \$513.00  &  \$501.00  &  \$505.00  &  \$505.00  &  \$491.00 \\\hline
Item 6  &  \$725.00  &  \$738.00  &  \$750.00  &  \$744.00  &  \$761.00 \\\hline\hline
Total &  \$3,000.00  &  \$2,910.00  &  \$3,090.00  &  \$3,030.00  &  \$3,000.00\\\hline
\else
Item 1 & \$352 & \$295 & \$395 & \$368 & \$324 \\\hline
Item 2 & \$98 & \$102 & \$98 & \$95 & \$105 \\\hline
Item 3 & \$460 & \$449 & \$510 & \$501 & \$476 \\\hline
Item 4 & \$852 & \$825 & \$832 & \$817 & \$843 \\\hline
Item 5 & \$513 & \$501 & \$505 & \$505 & \$491 \\\hline
Item 6 & \$725 & \$738 & \$750 & \$744 & \$761 \\\hline\hline
Total &  &  &  &  & \\\hline
Fair Share &  &  &  &  & \\\hline \fi

\end{tabular}

\end{center}
\begin{enumerate}
	\item Describe the first settlement of this estate and compute the surplus cash. 
	\ifsolns
	\$130 
	
	\begin{tabular}{lrccc}
 & Items & Fair Share & Initial Cash & Final Cash\\\hline
$A$   &  4,5  &   \$600.00   &   \$(765.00)  & (\$739.00)	\\\hline
 $B$   &    &   \$582.00   &   \$582.00   & \$608.00 	\\\hline
 $C$   &  1,3  &   \$618.00   &   \$(287.00)  & (\$261.00)	\\\hline
 $D$   &    &   \$606.00   &   \$606.00   & \$632.00 	\\\hline
 $E$  &  2,6  &   \$600.00   &   \$(266.00)  & (\$240.00)	\\\hline
\end{tabular} \else
	\fillwithlines{\stretch{1}}
\fi
	

	\item Describe the final settlement of this estate.
	\fillwithlines{\stretch{1}}
	

\end{enumerate}

\end{enumerate}

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\HOMEWORK
%</HWHEADER>

%<*HOMEWORK>


\begin{Denumerate}
  \item Aden, Ben, and Charles are dividing four pieces of furniture using the method of sealed bids.  Their bids on each of the items are given in the following table.
  
\begin{center}
	  \begin{tabular}{cccc}
 & Aden & Ben & Charles \\\hline
 Dresser & \$150 & \$300 & \$275 \\\hline
 
 Desk & \$180 & \$150 & \$165 \\\hline
 
 TV Cabinet & \$170 & \$200 & \$260 \\\hline
 Tapestry & \$400 & \$250 & \$500 \\\hline
\end{tabular}

\end{center}
		\begin{enumerate}
			\item Describe the first settlement of the items and compute the Surplus.
			\solution*{ Aden gets the Desk and \$120, Ben gets the Dresser, Charles gets the TV Cabinet and the Tapestry, but pays in \$360.  There is a surplus of \$240.
                }\vfill
			\item Describe the final settlement of the items.\vfill
				\solution*{ Aden gets the Desk and \$200, Ben gets the Dresser and \$80, Charles gets the TV Cabinet and the Tapestry, but pays \$280.
                }\vfill
			\end{enumerate}
	\item Kathy, Susie, and Judy are equal partners in a small bookstore.  They can't get along anymore, but they don't want to  sell the bookstore to an outsider so they decide to settle the matter using the method of sealed bids. Kathy bids \$240,000 for the business, Susie bids \$210,000, and Judy bids \$225,000.  Describe the final settlement of this bookstore.
			\solution{As the highest bidder, Kathy receives the store and owes \$155,000, Susie receives \$ 75,000  and Judy receives \$ 80,000. 
			}\vfill
\hwnewpage
		\item Three women (Carrie, Jeri, and Violet) share an apartment and wish to divide the chores: bathrooms, cooking, dishes, laundry, and vacuuming.  For each chore, they privately write the least they are willing to receive monthly (their negative valuation) in return for doing that chore.  The results are shown in the following table.
		
		\begin{center}
	\begin{tabular}{lccc}
 & Carrie & Jeri & Violet \\\hline
 Clean bathrooms & \$-20 & \$-30 & \$-40 \\\hline
 Do cooking & \$-50 & \$-10 & \$-25 \\\hline
 Wash dishes & \$-30 & \$-20 & \$-15 \\\hline
 Mow the lawn & \$-30 & \$-20 & \$-10 \\\hline
 Vacuum and dust & \$-20 & \$-40 & \$-15 \\\hline 
\end{tabular}
\end{center}
Divide the chores using the method of sealed bids.  Who does which chores?  Who gets paid, and how much? Who pays, and how much? 
		\solution*{Carrie cleans the bathrooms and pays \$12.  Jeri does the cooking and pays \$12.  Violet washes the dishes, mows the lawn, vacuums and dusts and gets paid \$23 (or \$24 with rounding).
		}

\end{Denumerate} \ENDHOMEWORK
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\solnstrue
\clearpage