\section{Simple Interest} \label{sec:SimpleInterest}
\index{interest!simple}
%\noindent
%In the following problems, there are no discounts for long-term rentals.
In this chapter, we retain the following conventions:
\begin{center}
	\begin{tabular}{l@{ is }l}
	1 year & 12 months\\
	1 year & 52 weeks \\
	1 year & 365 days \\
	1 week & 7 days \\
	1 day & 24 hours\\
	\end{tabular}
\end{center}
All interest rates will be annual unless specified otherwise.

\begin{enumerate}

  \item You rent a post-hole digger from RentAll to build a fence.
        The cost is \$12.95 per day, and you keep it for three days.
        How much will you pay? \solution{\$38.85}
          \vfill

  \item You're visiting Kentucky for work and need to rent a furnished apartment.
        The rate is \$100 per week.  You stay 18 days.  How much will you pay?
  \solution{\$257.14}
          \vfill

  \item You lease a car for \$2500 per year.
        How much will you pay if you keep the car for
        \begin{enumerate}
          \item 2 years? \solution{\$5,000.00}
                \vfill
          \item 5 months? \solution{\$1,041.67}
                \vfill
          \item 18 days? \solution{\$123.29}
                \vfill
          \item 6 weeks? \solution{\$288.46}
                \vfill
        \end{enumerate}

\clearpage

  %\item \fbox{\raisebox{0pt}[\height][1in]{\makebox[\textwidth][l]{\textbf{Simple Interest Formula:} \ifsolns $A=P*(1+r*t)$\fi}}}

  \item Guido ``the Organist'' Landini is a small-time loan shark in downtown Metropolis.
        He charges clients 43\% simple interest per year.
        What percentage interest does he charge on a loan of
        \begin{enumerate}
          \item 3 years? \solution{129\%}
                \fillwithlines{\stretch{1}}
          \item 1 month? \solution{3.58\%}
                \fillwithlines{\stretch{1}}
          \item 1 week? \solution{0.827\%}
                \fillwithlines{\stretch{1}}
          \item 3 weeks? \solution{2.48\%}
                \fillwithlines{\stretch{1}}
          \item 4 days? \solution{0.471\%}
                \fillwithlines{\stretch{1}}
        \end{enumerate}
  \item The interest charged is\ldots \index{interest}\ifsolns the interest rate as a decimal times the time (in the appropriate units) times the principle.\fi
        
        \fillwithlines{\stretch{1}}
\clearpage
  \item Loans: \index{loan}
        \ifsolns Loan as renting money\fi
        \fillwithlines{\stretch{2}}
        
  \item Calculations:
        \ifsolns Calculating simple interest \fi
        \fillwithlines{\stretch{1}}
  \item \boxedblank{\textbf{Simple Interest Formula:}\ifsolns $F=P(1+rt)$\fi} \index{interest!simple}

\clearpage

  \item Your cousin Ben borrows \$1800 from you for 3 years, at an annual simple interest rate of 7.0\%.
        How much will he repay you? %Solve for A, an odd multiple of years.
				\solution{\$2178}

        \vfill

  \item Suppose you borrow \$1000 for 4 months at an annual simple interest rate of 8.0\%.
        Find the total amount that you will have to repay. \solution{\$1026.66}
%Solve for A, less than a year.
        \vfill

  \item Suppose Fred borrows \$1500 from you for a period of 2 years, at an annual simple interest rate of 6.0\%.
        Find the amount of interest you will earn from Fred.\solution{\$180}
 %Solve for total interest paid.
        \vfill

\clearpage

  \item Joe took out a loan for 6 months at an annual simple interest rate of 4.0\%.
        The total amount he paid for the loan was \$1890.46.  How much did he borrow?     \solution{\$1853.39}
  % Solve for P.
        \vfill

  \item Suppose \$2000 is borrowed for 7 months, at the end of which \$2100 is repaid.
        Find the annual simple interest rate $r$. \solution{8.57\%}
  %Solve for r.
        \vfill
        
  \item Steve got a loan from the city to help repair the sidewalk in front of his home.
        He borrowed \$1800 at an annual simple interest rate of 5\%.
        The amount he repaid was \$2025.
        How long was his loan?      \solution{2.5 years, or 30 months}                % Solve for t.
        \vfill

\end{enumerate}

%</WORKSHEETS>

%<*HWHEADER>
\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\HOMEWORK
%</HWHEADER>

%<*HOMEWORK>


\begin{Fenumerate}

  \item Given the principal $P$, the annual interest rate $r$, and the time $t$,
        find the amount $F$ that must be repaid.
        \begin{enumerate}
          \def\prt#1#2#3{$P={}$\$#1, $r=#2\%$, $t={}$#3}
          \item \prt{15,000}{6}{5 years}
                \ifsolns \soln \fbox{\$19,500.00} \fi
                \studentsoln{\$19,500.00}
          \item \prt{5,300}{2}{3 months}
                \ifsolns \soln \fbox{\$5,326.50} \fi
          \item \prt{9,000}{4.5}{50 days}
                \ifsolns \soln \fbox{\$9,055.48} \fi
                \studentsoln{\$9,055.48}
        \end{enumerate}\vfill
				
  \item Of the four values $F$, $P$, $r$, and $t$ in the Simple Interest Formula,
        three are given to you.  Use the Simple Interest Formula to find the fourth one.
        \begin{enumerate}
          \item $F={}$\$12,000, $r=3\%$, $t={}$3 years
                \ifsolns \soln \fbox{\$11,009.17} \fi
                \studentsoln{\$11,009.17}
          \item $F={}$\$8,500,           $t={}$6 months, $P={}$\$8,200
                \ifsolns \soln \fbox{7.317\%} \fi
                \studentsoln{7.317\%}
          \item $F={}$\$4,250,  $r=7\%$,                 $P={}$\$3,500
                \ifsolns \soln \fbox{3.06 years} \fi
                \studentsoln{3.06 years}
        \end{enumerate}\vfill
  \item You borrow \$2,000 from the city to pay for sidewalk repairs.
        You promise to repay the loan in three years at 5\% simple interest.
        How much will you pay the city then?
        \ifsolns \par \soln \fbox{\$2,300} \fi
        \studentsoln{\$2,300} \vfill
\hwnewpage
  \item Istv\'an Nagy knows he will inherit \$10,000 from his dying uncle
        within five months.
        The bank will lend him money at 4\% simple interest.
        What is the largest amount of money he could borrow now,
        if he plans to use his inheritance to repay it in five months?
        \ifsolns \par \soln \fbox{\$9,836.07} \fi
        \studentsoln{\$9,836.07}\vfill
  \item Steve's friend Pierre wants to borrow \$220 from him for three months.
        If Steve wants to earn \$20 in interest on the loan,
        what percent simple interest should he charge?
        \ifsolns \par \soln \fbox{36.36\%} \fi
        \studentsoln{36.36\%}\vfill
  \item Ali has borrowed \$800 from his friend Jennifer, who is charging him 2\% simple interest.
        Eventually Ali repaid the loan, but it cost him \$200 to do so.
        For how long did Ali borrow Jennifer's money?
        \ifsolns \par \soln \fbox{12.5 years} \fi
        \studentsoln{12.5 years}\vfill

\end{Fenumerate} \ENDHOMEWORK

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage
